Efficient spherical Convolutional Neural Networks with Healpix sampling for cosmological applications

### 2018-05-25 scnn.tex
first pass at intro

	\todo{(Michaël) I would not invent a new term, SCNN, but rather say that convolutional neural networks on graphs (or GCNs) can be efficiently applied to (spherical?) cosmological applications.
	Arguments: the method is not new and hence don't deserve a new name. Emphasize that it is generic to the data structure, and that the sphere is simply a particular graph.
	Ideas: Graph Convolutional Networks for efficient spherical ???, Efficient spherical ??? with GCNs and Healpix sampling
}

@@

a) Cosmology has a lot of spherical data. Our method is simple and easy to use. Moreover, it is based on the widely used healpix sampling.
b) It is the most efficient spherical convolution\footnote{\todo{provably? cannot be faster than O(n) without approximations, e.g. sketching}}, requiring only $O(n)$ operations, where $n$ is the number of points.

\subsection{Potential applications}
	\assign{Tomek, Nathanaël, Michaël}

The analysis of spherical cosmological data, such as the cosmic microwave background \cite{...}, as done in \cite{he2018analysis}, is the target application of our method.

While our method was developed with cosmology in mind, it can easily target any problem where data live on a sphere. Examples include, but are certainly not limited to, (i) efficient compression and decompression of \ang{360} videos (see \cite{su2017learning}), (ii) \todo{data analysis on planets? (climate, forecasting, temperature, wind)}, (iii) \todo{particle physics? (jets on detectors, but they are usually cylindrical)}, (iv) \todo{applications in Cohen's papers?}.

Finally, not that those neural networks are not restricted to the sphere and can be applied to any problem where we have data on a graph, such as social, biological or infrastructure networks [some citations, e.g. brain Alzeihmer, particle physics, computer graphics].
% the convolution is not restricted to the sphere, the coarsening/pooling is

\todo{
\subsection{Healpix sampling}
Should we have this subsection to explain the advantages of Healpix?
\begin{itemize}
	\item Define the spherical grid?
	\item reference the "cubed sphere"~\cite{ronchi1996cubed}?
\end{itemize}
}

\begin{figure*}[!ht]
\centering
\includegraphics[width=0.7\textwidth]{figures/healpix-3layers.jpg}
\caption{3 smaller sclales of the healpix sampling.}
\label{fig:healpix_sampling}
\end{figure*}

@@

\section{Method}
% \begin{itemize}
% 	\item We build a graph using the healpix sampling
% 	\item Define Fourier transform and show that the harmonics are visually close to the spherical harmonics
% 	\item Define spherical convolution using the graph Fourier transform and show heat diffusion example
% 	\item Show the limits of the approach and explain why we cannot have a perfect spherical convolution with this technique
% \end{itemize}

@@

\section{Experiments}

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=0.45\textwidth]{figures/smooth_map_class_1.pdf}
\includegraphics[width=0.45\textwidth]{figures/smooth_map_class_2.pdf}
\caption{One sample of each class.}
\label{fig:map_sample}
\end{figure}

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=0.32\textwidth]{figures/result_order1.pdf}
\includegraphics[width=0.32\textwidth]{figures/result_order2.pdf}
\includegraphics[width=0.32\textwidth]{figures/result_order4.pdf}
\caption{Classification errors for the 3 different problems. \nati{TODO: order is not correct} }
\label{fig:results}
\end{figure}

@@

\section{Conclusion}
\assign{Nathanaël, Tomek, Michaël}

>>>>

	\todo{(Michaël) I would not invent a new term, SCNN, but rather say that convolutional neural networks on graphs (or GCNs) can be efficiently applied to (spherical?) cosmological applications.
	Arguments: the method is not new and hence don't deserve a new name. Emphasize that it is generic to the data structure, and that the sphere is simply a particular graph.
	Ideas: Graph Convolutional Networks for efficient spherical ???, Efficient spherical ??? with GCNs and Healpix sampling
    We make the code publicly available.
}

@@

Cosmological and astrophysical data often comes in a form of spherical sky maps.
Observables that cover large parts of the sky, such as the Cosmic Microwave Background (CMB) \citep{planck2015cosmologicalparameters,komatsu2011sevenyear,staggs2018recentdiscoveries}, neutral hydrogen \citep{santos2015cosmologySKA,HI4PI2016fullskyHI}, galaxy clustering \citep{alam2017clusteringgalaxies}, gravitational lensing \citep{troxel2017darkenergy,hildebrandt2017kidscosmological}, and others, have been used to constrain cosmological and astrophysical models. Deep Learning methods have recently started to gain popularity as an analysis tool in cosmology \citep{schmelze2017cosmologicalmodel,luciesmith2018machinelearning,gupta2018nongaussianinformation,gillet2018deeplearning,hassan2018reionizationmodels,aragoncalvo2018classyfyinglarge,ciuca2017cnnstring}.
Convolutional Neural Networks (CNNs) are particularly well suited for analysis of astrophysical data due to their ability to capture complicated, non-linear patterns, which are often present in the data.

So far these algorithms have been demonstrated on the flat images.
Deep learning-based analysis of larger survey area form requires dividing the data into smaller chunks and projecting them on a flat, 2D surface.
While this process can work efficiently as already demonstrated, it may be more convenient do work with deep learning directly on the sphere.
The most commonly used scheme in astrophysics to divide a sphere into pixels is Healpix \citep{gorski2005healpix}, which uses isolatitude sampling and pixels with the same area.

In this work we present an algorithm and software package \pkg{SCNN-temp-name}\footnote{github.com/SwissDataScienceCenter/scnn} which implements efficient deep convolutional neural networks on Healpix maps.
To achieve this, we represent the sphere $S^2$ as a graph of connected pixels, which forms a 2D manifold embedded in 3D space $\mathbb{R}^3$.
This approach is based on CNN formulations on graphs described by \citet{defferrard2016convolutional} and runs using the \pkg{TensorFlow} \citep{abadi2016tensorflow} engine.
The flexibility of the graph approach allows for faster analysis when data spans only a part of the sphere.
It is also possible to use multiple Healpix maps within our framework and use datasets in form of ``shells''.
These shells can span the radial direction, so that a tomographic analysis can be performed, or different frequencies, in case of data from observations in radio frequencies.

Formulations of deep learning on the sphere have been proposed by ~\citet{cohen2017convolutional,cohen2018spherical}.
In that method, a different sphere pixelisation scheme is used: the sphere pixels are mapped to flat space (??) and then generalized Fast Fourier Transform (FFT) is used for convolutions.
Other recently developed sampling methods include \citep{mcewen2011novelsampling}, where exact sampling theorem for equiangular MW sampling was presented.
In our work we focused on the Healpix sampling, as it is the most commonly used in cosmology.
It can be, however, extended to other sampling schemes, by straightforward change in definition of the graph.

Another advantage of our method is the speed of the convolution operation.
It is the most efficient spherical convolution\footnote{\todo{provably? cannot be faster than O(n) without approximations, e.g. sketching}}, requiring only $O(n)$ operations, where $n$ is the number of points.
MORE HERE

The method presented here restricts the shape of the filters learned by the network to be radial.
While this may be a less general design compared to the one used by conventional CNNs on flat images, we found it to still work efficiently for our test cases.
The layered structure of the network can partially compensate for that.
Further generalisations to non-radial filters may be possible [cite?], but we leave it to future work.

We give a practical demonstration of the application of our package to cosmological model discrimination using convergence maps, similar to \citep{schmelze2017cosmologicalmodel}.
In a simplified scenario, we classify convergence maps on parts of a sphere into two cosmological models.
These models were designed to have the same power spectrum in range $\ell < 1000$.
We compare the performance of the spherical CNN to a baseline algorithm, a SVM classifier which takes pixel histograms of these maps as input.
The comparison is made as a function of noise level on convergence maps and area of the sphere used.

The paper is organised as follows.
In Section \ref{sec:related} we describe existing approaches to convolutions on the sphere, graphs, and manifolds.
Section \ref{sec:method} we show the way to construct a graph using Healpix sampling and define the convolution operation.
The comparison between the spherical CNN and the benchmark method using the weak lensing mass map classification is presented in Section \ref{sec:experiments}.
We conclude in Section \ref{sec:conclusion}.
% Cite the following papers:
% \begin{itemize}
%     \item Healpix paper \citep{gorski2005healpix}
%     \item First DES mass maps \citep{chang2017curvedsky}
%     \item Healpix convolutions with asymetric beams \citep{mitra2011fastpixel}
%     \item Mass mapping on the sphere \citep{wallis2017mappingdark}
%     \item Wavelets on the sphere \citep{leistedt2016wavelet} and ball \citep{leistedt2012exactwavelets}
%     \item Planck main result \citep{planck2015cosmologicalparameters,komatsu2011sevenyear}
%     \item HI map \citep{HI4PI2016fullskyHI,}
%     \item more radio maps from Adam+Hamsa \citep{santos2015cosmologySKA}
%     \item more astro science on the sphere, what other survey make maps?
%     \item alternative sampling on the sphere \citep{mcewen2011novelsampling} MW sampling
%     \item cite kids tomographic power spectrum \citep{koehlinger2017kidstomographic}
% \end{itemize}

% a) Cosmology has a lot of spherical data. Our method is simple and easy to use. Moreover, it is based on the widely used healpix sampling.
% b) It is the most efficient spherical convolution\footnote{\todo{provably? cannot be faster than O(n) without approximations, e.g. sketching}}, requiring only $O(n)$ operations, where $n$ is the number of points.

% \subsection{Potential applications}
% 	\assign{Tomek, Nathanaël, Michaël}

% The analysis of spherical cosmological data, such as the cosmic microwave background \cite{...}, as done in \cite{he2018analysis}, is the target application of our method.

% While our method was developed with cosmology in mind, it can easily target any problem where data live on a sphere. Examples include, but are certainly not limited to, (i) efficient compression and decompression of \ang{360} videos (see \cite{su2017learning}), (ii) \todo{data analysis on planets? (climate, forecasting, temperature, wind)}, (iii) \todo{particle physics? (jets on detectors, but they are usually cylindrical)}, (iv) \todo{applications in Cohen's papers?}.

% Finally, not that those neural networks are not restricted to the sphere and can be applied to any problem where we have data on a graph, such as social, biological or infrastructure networks [some citations, e.g. brain Alzeihmer, particle physics, computer graphics].
% the convolution is not restricted to the sphere, the coarsening/pooling is

% \section{Healpix sampling}
% Should we have this subsection to explain the advantages of Healpix?
% \begin{itemize}
% 	\item Define the spherical grid?
% 	\item reference the "cubed sphere"~\cite{ronchi1996cubed}?
% \end{itemize}

% \begin{figure}[!ht]
% \centering
% \includegraphics[width=0.45\textwidth]{figures/healpix-3layers.jpg}
% \caption{3 smaller sclales of the healpix sampling.}
% \label{fig:healpix_sampling}
% \end{figure}

@@

\section{Method}
\label{sec:method}
% \begin{itemize}
% 	\item We build a graph using the healpix sampling
% 	\item Define Fourier transform and show that the harmonics are visually close to the spherical harmonics
% 	\item Define spherical convolution using the graph Fourier transform and show heat diffusion example
% 	\item Show the limits of the approach and explain why we cannot have a perfect spherical convolution with this technique
% \end{itemize}

@@

\section{Experiments}
\label{sec:experiments}

@@

\begin{figure*}[!ht]
\centering
\includegraphics[width=0.95\textwidth]{figures/smooth_map_class_1.pdf}
\includegraphics[width=0.95\textwidth]{figures/smooth_map_class_2.pdf}
\caption{One sample of each class.}
\label{fig:map_sample}
\end{figure*}

@@

\begin{figure*}[!ht]
\centering
\includegraphics[width=0.32\textwidth]{figures/result_order1.pdf}
\includegraphics[width=0.32\textwidth]{figures/result_order2.pdf}
\includegraphics[width=0.32\textwidth]{figures/result_order4.pdf}
\caption{Classification errors for the 3 different problems. \nati{TODO: order is not correct} }
\label{fig:results}
\end{figure*}

@@

\section{Conclusion}
\label{sec:conclusion}
\assign{Nathanaël, Tomek, Michaël}
